Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-10 Reviews & Rebuttal.md

## Reviewer 3

### Rating
- Abstract and Introduction are adequate?: Too long
- Conclusions/Future Work are convincing?: No
- Figures are adequate ? in number and quality: Yes
- Improve critical discussion ? validation: No
- Improve English?: No
- Needs comparative evaluation?: No
- Needs more experimental results?: No
- Originality Newness of the ideas expressed: 3
- Overall Rating Weighted value of above items: 3
- Paper formatting needs adjustment?: Yes
- Presentation Structure/Length/English: 2
- References are up-to-date and appropriate?: Yes
- Relevance Paper fits one or more of the topic areas?: 6
- Significance Is the problem worth the given attention?: 3
- Technical Quality Theoretical soundness/methodology: 3

### Observations for Author
This paper is an extension of the authors' previous VMV paper. The IVA appears quite the same with the addition of a decision tree that the authors claim is data mining. The paper is very long and does not need to be. The authors should be specific about their contributions (the IVA is not new) and condense the paper to about 8 pages maximum. Be specific about how this is an extension of the VMV paper and what is new and different. Compare and contrast with that paper rather than obfuscate the two. The future work has a paragraph ending in '(' I guess completing the section is future work? I would argue for a weak accept of this paper though it needs a rewrite to be publishable.

Rebuttal & Coverletter
======================

This rebuttal/coverletter is written in markdown. A graphical summary of the reviews can be found [here](https://docs.google.com/spreadsheets/d/1DTPt5eGe6PpMiNh38p4YsQrKLI3bqKhylZQKZKiZg4w/edit?usp=sharing).

@@

General Changes
---------------

- substantial reductions had to take place, since we were given a `short paper` slot with an 8-page limit.
	- redundancy between `Introduction` and `Epidemiological Background` was minimized as much as possible
	- `Related Work` section was shortened and covers less papers to meet the page limit.
		- as requested by [Reviewer 3], we elaborate on the differences towards the VMV paper in the `Related Work` section
	- `The Lumbar Spine Data Set` section was shortened, `Figure 2` was removed since we found it did not support the train of thought as we intended
	- `Experiments and Preliminary Results` section was shortened and changed due to the Reviewer feedback. It should be more precise and comprehensible.
- `Results` section now covers *all subjects*, subjects categorized by *gender* and *body mass index*
	- `Figure 3` (former `Figure 4`) was adapted accordingly
- `Summary & Conclusion` was completely revised and is now structured properly
	- due to the lack of time until the camera ready deadline as well as the page limit, we are not able to conduct a detailed user test. We did, however, discuss our approach and its drawbacks in more detail in the last section
- Figure 1 was completely removed since it does not supports the train of thought as we intended and due to lack of space

Reviewer 1
----------

- *"The exploration strategy is, in my opinion, not clear: there are not clear goals or net steps."*
	- role of Decision Trees and why they are chosen is now captured in the introduction of Section 6 - Interactive Decision Tree Visualization.
	- experiments and Preliminary Results completely restructured to make it comprehensible
- *"What is the role of decision trees explaining how numerical variables explains patient-based attributes in explaining the source of lower back pain?"*
	- the first step is always finding associations with back pain. As we conclude, that no such associations could be extracted at the given model precision, the focus shifts towards assessing, which variables can be described through the shape variables.
- *"Where is the interactive part of the images? Just browsing on dots on scatterplots and getting the trees?"*
	- Browsing the decision tree results is one interactive part. More important, however, is the ability to select target variables, which are used to subdivide the subjects, to assess their influence towards the result. Also, depending on how many subjects are included in a certain sub group (e.g. people suffering from a rare condition), the minimum subject count for a classification and the maximum error can be adjusted.
- *"Weak or no clue is provided about how images allowed for getting the final conclusions. How do the authors get the insights from the Figure 4? Focusing on the origin zone, with low error rate and little size tree?"*
	- We clarified this in the Figure caption.
- response to Minor Comments:
	- *"GPLOM Analysis Section. “The combination of the image variables with back pain is visualized as histogram at the left side…” left should be bottom"*: Back pain is encoded with presence (red) and absence (turquoise) in all plots left of the matrix diagonal. We clarified this in the text.
- *"The prototype is not available at blind.dnsalias.com"*
	- Regarding to our server logs we had a 100% uptime through the time of the reviewing process, we cannot comprehend whats gone wrong there

@@

Reviewer 2
----------

- *"the explanation of the domain is probably too lengthy"*
	- we removed redundancies and shortened the `Epidemiological Background` section
- *"the approach was not evaluated with user tests or at least with some inspection methods"*
	- due to the lack of time until the camera ready deadline as well as the page limit, we are not able to conduct a detailed user test. We did, however, discuss our approach and its drawbacks in more detail in the last section
- *"the paper would be strengthened if we could see some illustration where the proposed approach is uniquely capable of providing some actionable insights as compared to the existing approaches"*
	- the standard workflow of the epidemiologists is deriving hypotheses from the clinical day to day practice. The data are analyzed towards 2-3 independent variables at maximum towards a dependent variable, such as back pain. To our knowledge, there are no existing approaches used by epidemiologists in the day to day practice allowing for an explorative analysis using decision trees, so it is difficult to conduct a comparison as suggested. IVA methods amplify *hypotheses generation* and therefore aim to provide insight into unexpected relationships in the data.

### 2014-12-10 draft.tex
revised first three sections, including an error in Figure 1

\item The detection of associations between image-based, socio-demographic and medical variables for hypothesis generation,

@@

(1) Physicians formulate hypotheses based on observations made in their clinical practice; (2) Epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) Statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.

@@

\emph{Hypothesis generation} is amplified by concurrently analyzing heterogenous variables at once using correlation measures.

@@

This approach aims to trigger \emph{hypothesis generation} by outlining tendencies between these variables.

@@

In a prior work, we analyzed the lumbar spine variability of 490 \texttt{SHIP} subjects \cite{Klemm2013VMV}.

@@

To classify breast tumors, they train a decision-tree classifier, concluding that the extracted kinetic and morphological features alone are not sufficient for tumor type classification.

@@

\cite{Niemann2014} assess risk factors for hepatic steatosis (fatty liver disease) risk factors using decision trees.

@@

They present an interactive data mining tool, that can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.

@@

Unique in our work is the combination of classification techniques with an IVA approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.

@@

These observations brought us to the decision to incorporate classification techniques to assess the influence of the image-derived variables.
%%
% \section{\uppercase{Evaluation of Decision Trees}}
\section{\uppercase{Interactive Decision Tree Visualization}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variables explain non-image variables.
%%
%The presented method aims to support an explorative analysis by trying to identify relationships with all non-image variables and also 
%The nature of this analysis is more explorative as it includes all 
%%
Decision trees are used to create predictive models.
%%
These models are build w.r.t. all input variables and capture more complex relationships than correlation coefficients.
%%
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
%They work with numerical as well as categorical data.
%%
Readability is only granted for small trees.
%%
%Large, complex trees are difficult to trace and to understand.
%%
Too many branches impose overfitting to the data \cite{DecisionTree}.
%%
\\\\
\noindent \textbf{Creating Decision Trees.}
We use the C4.5 algorithm, which builds decision trees based on information entropy on a data set.
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%
%\subsection{The C4.5/C5.0 Algorithm}
%\label{sec:C5.0}
%The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
%Such a calculation requires a numeric or categorical target variable.
%%
% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
% \begin{algorithm}[]
%  Check for base cases\;
%  \For{each attribute $a$}{
%   Find the normalized information gain ratio from splitting on $a$\;
%  }
%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
%  Create a decision node that splits on $a_{best}$\;
%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
%  \caption{Building a decision tree using the C4.5 Algorithm}
%  \label{alg:c4.5}
% \end{algorithm}
%C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
%We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
% Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
% %%
% Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
% %%
% In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%\subsection{Visualizing Decision Trees}

@@

Subjects experiencing \emph{pain in the last seven days} can also be predicted for this group using the same features with a tree consisting of $8$ rules and has an error of $0.35$

@@

Methods supporting \emph{hypothesis generation} based on image information are new to the application domain.

>>>>

\item The detection of associations between image-based, socio-demographic and medical variables for \emph{hypotheses generation},

@@

(1) physicians formulate hypotheses based on observations made in their clinical practice; (2) epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.

@@

\emph{Hypotheses generation} is amplified by concurrently analyzing heterogenous variables at once using correlation measures.

@@

This approach aims to trigger \emph{hypotheses generation} by outlining tendencies between these variables.

@@

In a prior work, we analyzed the lumbar spine variability of 490 subjects \cite{Klemm2013VMV}.

@@

To classify breast tumors, they train a decision-tree classifier, concluding that the extracted kinetic and morphological variables alone are not sufficient for tumor type classification.

@@

\cite{Niemann2014} assess risk factors for hepatic steatosis (fatty liver disease) using decision trees.

@@

Their interactive data mining tool can analyze association rules and highlight relations.
%They present an interactive data mining tool, that can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.

@@

Unique in our work is the combination of classification techniques with an IVA approach by observing interesting variable relations in the context of image-derived variables using multiple decision trees for one visualization.

@@

These observations brought us to the decision to incorporate classification techniques to assess the influence of the image-derived variables.
%%
% \section{\uppercase{Evaluation of Decision Trees}}
\section{\uppercase{Interactive Decision Tree Visualization}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variables explain non-image variables.
%%
%The presented method aims to support an explorative analysis by trying to identify relationships with all non-image variables and also 
%The nature of this analysis is more explorative as it includes all 
%%
Decision trees are used to create predictive models.
%%
These models are build w.r.t. all input variables and capture more complex relationships than correlation coefficients.
%%
Leafs of a decision tree represent class labels, branches represent variable conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
%They work with numerical as well as categorical data.
%%
Readability is only granted for small trees.
%%
%Large, complex trees are difficult to trace and to understand.
%%
Too many branches impose overfitting to the data \cite{DecisionTree}.
%%
\\\\
\noindent \textbf{Creating Decision Trees.}
We use the C4.5 algorithm, which builds decision trees based on information entropy on a data set.
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%
%\subsection{The C4.5/C5.0 Algorithm}
%\label{sec:C5.0}
%The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
%Such a calculation requires a numeric or categorical target variable.
%%
% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
% \begin{algorithm}[]
%  Check for base cases\;
%  \For{each attribute $a$}{
%   Find the normalized information gain ratio from splitting on $a$\;
%  }
%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
%  Create a decision node that splits on $a_{best}$\;
%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
%  \caption{Building a decision tree using the C4.5 Algorithm}
%  \label{alg:c4.5}
% \end{algorithm}
%C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
%We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
% Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
% %%
% Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
% %%
% In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%\subsection{Visualizing Decision Trees}

@@

Subjects experiencing \emph{pain in the last seven days} can also be predicted for this group using the same variables with a tree consisting of $8$ rules and has an error of $0.35$

@@

Methods supporting \emph{hypotheses generation} based on image information are new to the application domain.
